---
title: "Crosswalk methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Crosswalk methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(eq5dsuite)
```

## Introduction

Crosswalk methods allow EQ-5D values to be estimated when
a country-specific value set exists for one instrument
version but not another. `eq5dsuite` implements three
widely used approaches:

- **Original crosswalk** (`eqxw`) — maps EQ-5D-5L responses
  to EQ-5D-3L values (van Hout et al., 2012)
- **Reverse crosswalk** (`eqxwr`) — maps EQ-5D-3L responses
  to EQ-5D-5L values (van Hout and Shaw, 2021)
- **UK-specific crosswalk** (`eqxw_UK`) — NICE-recommended
  mapping for UK analyses (Hernandez Alava et al., 2023)

## When to use crosswalk methods

Use crosswalk methods when:

- You have EQ-5D-5L data but only a 3L value set exists
  for your country → use `eqxw()`
- You have EQ-5D-3L data but only a 5L value set exists
  → use `eqxwr()`
- You are conducting a NICE technology appraisal and have
  EQ-5D-5L data → use `eqxw_UK()` until a NICE-approved
  5L value set is available

## Original crosswalk: eqxw()

```{r eqxw-example}
eq5d5l_data <- data.frame(
  id = 1:5,
  mo = c(1, 2, 3, 1, 2),
  sc = c(1, 1, 2, 1, 3),
  ua = c(2, 1, 3, 1, 2),
  pd = c(2, 3, 2, 1, 4),
  ad = c(1, 2, 1, 3, 2)
)

# Apply crosswalk using UK 3L value set
eq5d5l_data$value_xw <- eqxw(
  eq5d5l_data,
  country   = "UK",
  dim.names = c("mo", "sc", "ua", "pd", "ad")
)

eq5d5l_data
```

## Reverse crosswalk: eqxwr()

```{r eqxwr-example}
eq5d3l_data <- data.frame(
  id = 1:5,
  mo = c(1, 2, 1, 3, 2),
  sc = c(1, 1, 2, 2, 1),
  ua = c(1, 2, 1, 3, 2),
  pd = c(2, 2, 1, 3, 3),
  ad = c(1, 1, 2, 2, 1)
)

# Apply reverse crosswalk using Spain 5L value set
eq5d3l_data$value_rxw <- eqxwr(
  eq5d3l_data,
  country   = "ES",
  dim.names = c("mo", "sc", "ua", "pd", "ad")
)

eq5d3l_data
```

## UK-specific crosswalk: eqxw_UK()

The UK-specific crosswalk requires age and sex in addition
to the five EQ-5D dimensions:

```{r eqxw-uk-example}
uk_data <- data.frame(
  id   = 1:5,
  mo   = c(1, 2, 1, 3, 2),
  sc   = c(1, 1, 2, 2, 1),
  ua   = c(1, 2, 1, 3, 2),
  pd   = c(2, 2, 1, 3, 3),
  ad   = c(1, 1, 2, 2, 1),
  age  = c(45, 62, 38, 71, 55),
  male = c(1,  0,  1,  0,  1)
)

uk_data$value_uk <- eqxw_UK(
  uk_data,
  age  = "age",
  male = "male"
)

uk_data
```

## Crosswalk compatibility with custom value sets

All crosswalk functions are compatible with custom value
sets added via `eqvs_add()`. The mapping and valuation
steps are independent, so any compatible value set can
be used without modification.

## References

van Hout B, et al. (2012). Interim scoring for the
EQ-5D-5L. *Value in Health*, 15(5), 708--715.

van Hout B, Shaw JW (2021). Mapping EQ-5D-3L to EQ-5D-5L.
*Value in Health*, 24(9), 1285--1293.

Hernandez Alava M, et al. (2023). Estimating the
relationship between EQ-5D-5L and EQ-5D-3L.
*PharmacoEconomics*, 41(2), 199--207.
